---
title: Metric Multidimensional Unfolding
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: We review Schönemann's technique to reconstruct a configuration of $n+m$   points from an $n\times m$ off-diagonal submatrix of the distance matrix. 
  The technique is meant for error-free distances, but we pat spacial attention 
  to what happens if it is applied to dissimilarities which may not have a 
  perfect representation in Euclidean space. Examples are analyzed and R 
  programs are provided. 
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
```

```{r somecode, echo = FALSE}
ei <- function(i, n) ifelse(i == 1:n, 1, 0)
jmat <- function(n) diag(n) - 1/n
matrixPrint <- function(x,
                        digits = 6,
                        width = 8,
                        format = "f",
                        flag = "+") {
  print(noquote(
    formatC(
      x,
      digits = digits,
      width = width,
      format = format,
      flag = flag
    )
  ))
}
source("schoenemann.R")
```

\sectionbreak 

**Note:** This is a working manuscript which can be expanded/updated
multiple times. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/schoenemann> 

\sectionbreak

# Introduction

At the core of classical metric Multidimensional Scaling (MDS) is the
Schoenberg-Young-Householder-Torgerson-Gower (SYHTG) Theorem. It was first proved in a distance geometry context by @schoenberg_35 and then (independently, it seems)  by @young_householder_38 for application in the very first versions of MDS in psychometrics. It was given its current more computational form by @torgerson_58 and @gower_66.

Suppose $D$ is a square, symmetric, and hollow non-negative matrix with dissimilarities between $n$ objects. Define 
\begin{equation}
C:=-\frac12 J_n(D\times D)J_n
\end{equation}
where $J_n=I-n^{-1}e_ne_n'$ is the centering matrix, $e_n$ has all its 
elements equal to one, and $D\times D$ is the elementwise square of $D$.
The SYHTG theorem says that there is an $n\times p$ matrix $X$ such that
\begin{equation}
d_{ij}=\sqrt{\sum_{s=1}^p(x_{is}-x_{js})^2}
\end{equation}
if and only if $C$ is positive semidefinite with rank $r\leq p$. Moreover
in that case $n\times r$ solutions $X$ can be found from any full rank decomposition $C=XX'$. 

The SYHTG result has an existence part, which says that a Euclidean
representation exists if and only if the matrix $C$ is positive 
semi-definite. This is an algebraic result, which is of limited
direct relevance for data analysis. In practical situations data
will be fallible, and we merely expect that dissimilarities can be 
approximated by Euclidean distances. The second part of the SYHTG
result is an approximation result, which says that a $p$-dimensional 
Euclidean configuration of points with distances that approximate
the given dissimilarities can be found by using the $p$ largest eigenvalues
of $C$ and the corresponding eigenvectors.

Multidimensional Unfolding (MDU) is a special case of MDS in which
there are two disjoint sets of objects. The first set has $n$
elements, the second set has $m$ elements. The non-negative matrix $D$ is
$n\times m$ and contains the dissimilarities between the elements
of the first and second set. The within-set dissimilarities are
missing. There were some tentative attempts to generalize the
SYHTG theorem to off-diagonal data of this form, but not much was
achieved until the work of @schoenemann_70, which also has a review of the accomplishments until 1970. 

Schönemann's technique is developed for perfect data, i.e. for dissimilarities that can be faithfully represented as distances in Euclidean space. It is
an exact algebraic method that does not involve any approximation, similar 
to the embedding part of the SYHTG theorem. In the second part of his paper
Schönemann applies the embedding calculations to "fallible" data, using
some small artificial examples. Unlike the SYHTG approximation part, which
has a clear least squares interpretation, these calculations proceed 
by what Joe Kruskal used to call "ignoring errors". 

@heiser_deleeuw_A_79 present various variations of the Schönemann technique,
and review the literature until 1979. They also compare the results on various examples with metric MDS results obtained from smacof, the standard iterative
least squares MDS technique. And they provide an example of applying their
Schönemann variations on a real data analysis example. Although the standard
MDS textbook of @borg_groenen_05 pays a lot of attention to unfolding, the work
of Schönemann is only mentioned in a single footnote.

# Perfect Data

Suppose $D$ is an $n\times m$ matrix with non-negative numbers. The problem we
study in this paper is solving the system of equations
\begin{equation}
d_{ij}=\sum_{s=1}^p(x_{is}-y_{js})^2.\label{eq-d}
\end{equation}
for an  $n\times p$ matrix $X$ and an $m\times p$ matrix $Y$.
In matrix notation \eqref{eq-d} becomes
\begin{equation}
D=\alpha e_m'+e_n\beta'-2XY',\label{eq-dmat}
\end{equation}
with $\alpha:=\text{diag}(XX')$ and $\beta:=\text{diag}(YY')$ and with $e_n$ and $e_m$ vectors of appropriate length with all elements equal to one. In this section we study the case in which we have "perfect data", i.e. in which \eqref{eq-dmat} has a solution $(X,Y)$ for some $p$. Our problem is to recover $X$ and $Y$ from $D$.

Suppose $J_n=I-n^{-1}e_ne_n'$ and $J_m=I-m^{-1}e_me_m'$ are centering matrices. Then
\begin{equation}
C:=-\frac12 J_nDJ_m=J_nXY'J_m.\label{eq-cdef}
\end{equation}
Suppose $C$ has rank $p$ and $C=GH'$ is any full-rank decomposition of $C$. Being able to construct $X$ and $Y$ from $C$ assumes that $J_nX$ and $J_mY$ have the same rank as $X$ and $Y$, which means that we assume that $e_n$ is not in the column-space of $X$ and $e_m$ is not in the column space of $Y$. If that assumption is violated the full-rank decomposition has rank $q<p$. In an extreme case, suppose $Y=e_my'$ for some $p$-vector $y$. Then all rows of $D$ are the same and $C=0$.

From \eqref{eq-cdef} it follows that there exist
$p\times p$ nonsingular matrices $T$ and $S$ and vectors $\mu$ and $\nu$ 
of length $p$ such that
\begin{subequations}
\begin{align}
X&=GT+e_n\mu',\\
Y&=HS+e_m\nu',\\
TS'&=I.
\end{align}
\end{subequations}
Consequently
\begin{subequations}
\begin{align}
XX'&=GMG'+GT\mu e_n'+e_n\mu'T'G'+\mu'\mu\ e_ne_n',\label{eq-5a}\\
YY'&=HNH'+HS\nu e_m'+e_m\nu'S'H'+\nu'\nu\ e_me_m',\label{eq-5b}\\
XY'&=GH'+GT\nu e_m'+e_n\mu'S'H'+\mu'\nu\ e_ne_m',\label{eq-5c}
\end{align}
\end{subequations}
where $M:=TT'$ and $N:=SS'$. Note that $N=M^{-1}$.

Define an $n\times \frac12 p(p+1)$ matrix $G\star G$ which has a column
for each pair of indices $1\leq s\leq t\leq p$. Columns of $G\star G$ are elementwise products of columns of $G$ according to
\begin{equation}
(G\star G)_{(s,t)}:=\begin{cases}2g_sg_t&\text{ if }s\neq t,\\
g_s^2&\text{ if }s=s.
\end{cases}
\end{equation}
Matrix $H\star H$ is defined in the same way, using the colums of $H$. Next define the $\frac12 p(p+1)$-element vectors $\overrightarrow{m}$ and $\overrightarrow{n}$, which have the 
diagonal and subdiagonal elements of $M$ and $N$, in the same order as the
columns of $G\star G$ and $H\star H$. Then $\text{diag}(GMG')=(G\star G)\overrightarrow{m}$ and
$\text{diag}(HNH')=(H\star H)\overrightarrow{n}$.

In addition to \eqref{eq-5a}-\eqref{eq-5c} we have
\begin{subequations}
\begin{align}
\alpha e_m'&=(G\star G)\overrightarrow{m} e_m'+2GT\mu e_m'+\mu'\mu e_ne_m',\label{eq-7a}\\
e_n\beta'&=e_n\overrightarrow{n}'(H\star H)'+2e_n\nu'S'H'+\nu'\nu e_ne_m'.\label{eq-7b}
\end{align}
\end{subequations}
Substituting \eqref{eq-5a}-\eqref{eq-5c} and \eqref{eq-7a}-\eqref{eq-7b} in equation \eqref{eq-dmat} gives
\begin{multline}
D=-2GH'+2GT\eta e_m'-2e_n\eta'S'H'+\\(G\star G)\overrightarrow{m} e_m'+e_n\overrightarrow{n}'(H\star H)'+\eta'\eta e_ne_m',\label{eq-symm}
\end{multline}
with $\eta:=\mu-\nu$.

We now take what could be described as a "row-oriented approach", by eliminating all references to $H$ and $S$ from equation \eqref{eq-symm}.
Define $F:=D+2GH'$ and $\tilde F:=J_nF$. Then, from \eqref{eq-symm},
\begin{equation}
\tilde F=2J_nGT\eta e_m'+J_n(G\star G)\overrightarrow{m} e_m'\label{eq-feq}
\end{equation}
Since $C=GH'$ it follows that $D+2GH'=D-J_nDJ_m$ and 
all $m$ columns of $\tilde F$ are identical to the centered row-averages of $D$.

Assume $\tilde F$ is non-zero, i.e. that not all row-sums of $D$ are the same. Take any of its columns, name it $f$. 
Define a matrix 
\begin{equation}
K:=\begin{bmatrix}2J_nG&J_n(G\star G)\end{bmatrix}.\label{eq-kdef}
\end{equation}
We must have
\begin{equation}
f=\begin{bmatrix}2J_nG&J_n(G\star G)\end{bmatrix}\begin{bmatrix}T\eta\\\overrightarrow{m}\end{bmatrix}.\label{eq-linsys}
\end{equation}
This is a system of $n$ linear equations in $\frac12 p(p+3)$ unknowns. Suppose it has a unique solution for $T\eta$ and $\overrightarrow{m}$. If $n\geq\frac12 p(p+3)$ then the solution can be computed by inverting any non-singular submatrix
of order $\frac12 p(p+3)$. No matter how we choose the submatrix, the solution will be the same. From the solution of \eqref{eq-linsys} use the elements
of $\overrightarrow{m}$ to reconstruct $M$. Use any decomposition method to find $T$ from $M=TT'$ and use $T$ to find $\eta$ and $S=(T')^{-1}$.
Assume without loss of generality that $\nu=0$. Then $Y=HS$ and $X=GT+e_n\mu'$
gives a solution, with $Y$ column-centered.

If the system of equations \eqref{eq-dmat} has a solution, then the linear equations \eqref{eq-linsys} are a consequence of \eqref{eq-dmat}, and thus have at least one solution with a positive definite matrix $M$. If \eqref{eq-linsys} has more than one solution, because the matrix $K$ has rank less than $\frac12p(p+3)$, then we must find a solution for which $M$ is positive definite. Conversely, if \eqref{eq-linsys} has a solution with $M$ positive definite then system \eqref{eq-dmat} has a solution with dimension $p=\text{rank}(C)$.

Note that it is also possible to go the "column-oriented" way, by defining 
$\tilde F$ as $FJ_m$, which now has identical rows depending on $N$ and $S$.
This also leads to a solution for $X$ and $Y$, which in the perfect case should be identical to the "row-oriented" one. We went the "row-oriented" way because we have the practical situations in mind in which $n$ is larger, often much larger, than $m$. And we can always find the "column-oriented" solution by applying the "row-oriented" calculations to the transpose of $D$.
For example, if all row-sums of $D$ are the same then $\tilde F=0$ and our "row-oriented" technique fails. In this case we can analyze the transpose of $D$ instead, which may have  non-constant row-sums.

\sectionbreak

# Artificial Example

This section has in-line R code and R output for a $9\times 3$ artificial example. In the code we use the same names (in lower case) for the key matrices as in our formulas (in upper case). The full-rank decomposition of $C$ is computed from its singular value decomposition. Note that the solutions, which are $\hat X$ and $\hat Y$ in the output, are not the same as $X$ and $Y$ used to generate the data because of the rotational and translational indeterminacy. But the distances $\hat D$ between $\hat X$ and $\hat Y$ are the same as the distances $D$ between $X$ and $Y$, which is what matters.

```{r arti}
x <- matrix(c(1:9, 1, 2, 3, 1, 2, 3, 1, 2, 3), 9, 2) / 10
y <- matrix(c(1, 1, -2, 1, 0, -1), 3, 2)
matrixPrint(x, digits = 2, width = 4)
matrixPrint(x, digits = 2, width = 4)
a <- diag(tcrossprod(x))
b <- diag(tcrossprod(y))
d <- outer(a, b, "+") - 2 * tcrossprod(x, y)
matrixPrint(d, digits = 4, width = 6)
c <- -(jmat(9) %*% d %*% jmat(3)) / 2
matrixPrint(c, digits = 4, width = 6)
sv <- svd(c)
e <- diag(sqrt(sv$d[1:2]))
g <- sv$u[, 1:2] %*% e
matrixPrint(g, digits = 4, width = 6)
h <- sv$v[, 1:2] %*% e
matrixPrint(h, digits = 4, width = 6)
gg <- cbind(g[, 1]^2, 2 * g[, 1] * g[, 2], g[, 2]^2)
hh <- cbind(h[, 1]^2, 2 * h[, 1] * h[, 2], h[, 2]^2)
f <- jmat(9) %*% (d + 2 * tcrossprod(g, h))
matrixPrint(f, digits = 4, width = 6)
jg <- jmat(9) %*% g
jgg <- jmat(9) %*% gg
k <- cbind(2 * jg, jgg)
fm <- apply(f, 1, mean)
cof <- qr.solve(k, fm)
m <- matrix(c(cof[3], cof[4], cof[4], cof[5]), 2, 2)
t <- t(chol(m))
mn <- solve(t, cof[1:2])
xhat <- g %*% t + matrix(mn , 9, 2, byrow = TRUE)
matrixPrint(xhat, digits = 4, width = 6)
s <- t(solve(t))
yhat <- h %*% s
matrixPrint(yhat, digits = 4, width = 6)
ahat <- diag(tcrossprod(xhat))
bhat <- diag(tcrossprod(yhat))
dhat <- outer(ahat, bhat, "+") - 2 * tcrossprod(xhat, yhat)
matrixPrint(dhat, digits = 4, width = 6)
```

\sectionbreak

# Imperfect Data

For real data the system \eqref{eq-dmat} will generally not have a perfect solution 
for any $p$ (i.e. there is no Euclidean configuration for which the $D$
are squared distances). At various steps our technique necessarily differs from the technique in the perfect case.

First we have to pick a $p$. This can be done a priori, for instance
by using the MDS default $p=2$. Or we can pick $p$ by eye-balling the
singular values of $C$ and try to find an "elbow" familiar from 
factor analysis. Instead of a full rank decomposition of $C$ we
then use a rank $p$ approximation $GH'$, for example by using the
$p$ dominant singular values and their singular vectors. 

Now proceed with the calculations as in the perfect case. For imperfect data
with $GH'\neq C$ the columns of $F$ from \eqref{eq-dmat} are not identical any
more, and not equal to the centered row-means. So picking different columns will give different results. In some cases it may make sense to choose $f$ equal to the row-averages of $F$. But note that, no matter how good or bad the fit of 
$GH'$ to $C$ is, the row-averages of $F$ are always equal to the row-averages of $D$. There are other options, as long as they give the correct result in the perfect case. Thus picking the row-wise median or minimum or maximum is also a 
possibility. Remember that a rule that makes $f$ small (in norm) will make $M$, and consequently $T$ and $X$, small and will make $S$ and $Y$ large. 

In some cases we can actually use the fact that the columns of $F$ are
different to perform $m$ analyses, using each column of $F$ in turn. The $m$ different results then give some information about the fit and the stability of
the solution. 

In the next step of the technique we may find that
system \eqref{eq-linsys} is overdetermined and not solvable, 
or singular with multiple solutions. The usual case will be 
that no exact solution exists. As a compromise in that case we 
can use the Moore-Penrose inverse to compute the
minimum norm least squares solution. 

And finally, and this is the most
troublesome problem, we may find that $M$ constructed from
the least squares solution is not positive definite.  In that case
case we cannot find a $T$ such that $M=TT'$ and we are stuck.
As this happens, for now, we give up.

Of course these problems will be less serious if the fit is good, i.e.
if the elements of $D$ are approximately squared Euclidean distances.

The only choice that could make a difference is the full-rank decomposition.
Suppose $C=\tilde G\tilde H'$ is a different one. It follows that
$\tilde G=GA$ and $\tilde H=HB$, with $AB'=I$. This does not change
$\tilde F$, but it does change the matrix $K$ from \eqref{eq-kdef}. 
If $T$ and $\eta$ solve ..., then $\tilde T=A^{-1}T$ and $\tilde\eta=A\eta$
solve. Thus in the case of perfect data it does not matter which
full-rank decomposition we choose.


\sectionbreak

# Real Examples

## Airline Distances

Our first real example is a matrix of airline distances between 18 cities. The
cities are
```{r airlineData, echo = FALSE}
source("airlineData.R")
print(as.matrix(row.names(airline), 18, 1), quote = FALSE)
```
We start by computing the metric MDS solution on the full matrix.
```{r airfull, echo = FALSE}
dd <- airline^2
da <- apply(dd, 1, mean)
dm <- mean(dd)
ca <- -(dd - outer(da, da, "+") + dm) / 2
ea <- eigen(ca)
xa <- ea$vectors[,1:2] %*% diag(sqrt(ea$values[1:2]))
```
```{r acomplot, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Airline Distances, Configuration Plot, Complete Data"}
par(pty="s")
plot(xa, type = "n", xlab = "dim 1", ylab = "dim 2")
text(xa, row.names(airline), col = "RED", cex = 1.5)
```
```{r acomplot2, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Airline Distances, Dist-Dhat Plot, Complete Data"}
plot(as.dist(airline), dist(xa), col = "BLUE", pch = 19, cex = 1.5,
     xlab = "D", ylab = "DHAT")
```
We now create a $9\times 9$ off-diagonal submatrix by using the even-numbered cities as rows and the rest as columns. 
```{r airect1, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Airline Distances, Configuration Plot, Off-diagonal Submatrix"}
par(pty="s")
hr <- schoenemann(airlineRect^2)
plot(rbind(hr$xhat, hr$yhat), type = "n", xlab = "dim 1", ylab = "dim 2")
  text(hr$xhat,
       rowLabels,
       col = "RED",
       cex = 1.5)
  text(hr$yhat,
       colLabels,
       col = "BLUE",
       cex = 1.5)
```
```{r airect2, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Airline Distances, Configuration Plot, Off-diagonal Submatrix, Transposed"}
hc <- schoenemann(t(airlineRect^2))
plot(rbind(hc$xhat, hc$yhat), type = "n", xlab = "dim 1", ylab = "dim 2")
  text(hc$xhat,
       colLabels,
       col = "RED",
       cex = 1.5)
  text(hc$yhat,
       rowLabels,
       col = "BLUE",
       cex = 1.5)
```

## Gold

Our next example is the same as the one in @heiser_deleeuw_A_79.
The data are taken from the *Power in the Classroom* study of @gold_58.
Children in elementary school, divided into eight groups, judge how important
17 characteristics of their classmates are for their power in the classroom. 
Data are the times each of properties were rated as "very important" by the
eight age-sex groups.

```{r gold, echo = FALSE}
gold <-
structure(list(V1 = c(13.5, 1, 13.5, 16.5, 12, 9.5, 2, 9.5, 5.5, 
5.5, 15, 3, 5.5, 11, 16.5, 8, 5.5), V2 = c(13, 17, 6.5, 3, 4, 
13, 15.5, 1, 5, 11, 13, 8.5, 6.5, 10, 15.5, 8.5, 2), V3 = c(17, 
13, 12, 14, 11, 15, 3, 4, 1, 7.5, 5, 9, 10, 7.5, 16, 6, 2), V4 = c(15L, 
6L, 15L, 17L, 15L, 13L, 3L, 11L, 4L, 1L, 2L, 10L, 5L, 12L, 7L, 
9L, 8L), V5 = c(16L, 6L, 13L, 17L, 14L, 12L, 3L, 9L, 4L, 7L, 
2L, 8L, 1L, 15L, 10L, 11L, 5L), V6 = c(17, 10, 12.5, 11, 15, 
16, 5, 6, 2, 1, 8, 9, 7, 14, 12.5, 3, 4), V7 = c(17L, 9L, 15L, 
13L, 16L, 14L, 2L, 6L, 1L, 4L, 5L, 11L, 8L, 10L, 12L, 7L, 3L), 
    V8 = c(16L, 4L, 14L, 13L, 12L, 15L, 2L, 9L, 5L, 7L, 3L, 6L, 
    1L, 17L, 10L, 11L, 8L)), class = "data.frame", row.names = c(NA, 
-17L))
goldMatrix <- as.matrix(gold)
goldLabels <- c("3MM","3MF","3FM","3FF","4MM","4MF","4FM","4FF")
hgold0 <- schoenemann(goldMatrix^2)
hgold1 <- schoenemann(goldMatrix^2, fsel = 1)
hgold2 <- schoenemann(goldMatrix^2, fsel = 2)
hgold3 <- schoenemann(goldMatrix^2, fsel = 3)
hgold4 <- schoenemann(goldMatrix^2, fsel = 4)
hgold5 <- schoenemann(goldMatrix^2, fsel = 5)
hgold6 <- schoenemann(goldMatrix^2, fsel = 6)
hgold7 <- schoenemann(goldMatrix^2, fsel = 7)
hgold8 <- schoenemann(goldMatrix^2, fsel = 8)
```
```{r jplot, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Gold Data, Configuration Plot"}
par(pty="s")
plot(rbind(hgold0$xhat, hgold0$yhat), type = "n", xlab = "dim 1", ylab = "dim 2")
text(hgold0$xhat, as.character(1:17), col = "BLUE", cex = 2)
text(hgold0$yhat, goldLabels, col = "RED", cex = 2)
```

```{r dplot, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Gold Data, Dist-Dhat Plot"}
par(pty="s")
plot(hgold0$d, hgold0$dhat, cex = 2, col = "BLUE", pch = 19, xlab = "D", ylab = "DHAT")
```


```{r yplot, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Gold Data, Column Configuration from 8 Column and One Average Analysis"}
par(pty="s")
plot(rbind(hgold1$y,hgold2$y,hgold3$y,hgold4$y,hgold5$y,hgold6$y,hgold7$y,hgold8$y),type="n", xlab = "dim 1", ylab = "dim 2")
text(hgold1$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold2$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold3$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold4$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold5$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold6$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold7$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold8$y, as.character(1:8), col = "BLUE", cex = 2)
text(hgold0$y, as.character(1:8), col = "RED", cex = 3)
```

## Roskam

In 1968 @roskam_68 asked 39 of the staff members of the Psychology Laboratory
at Nijmegen University to rank the nine different departments in the laboratory as to the importance for their work. The departments are

1. Social Psychology
2. Educational and Developmental Psychology
3. Clinical Psychology
4. Mathematical Psychology and Statistics
5. Experimental Psychology
6. Cultural Psychology and Psychology of Religion
7. Industrial Psychology
8. Test Construction and Validation
9. Physiological and Animal Psychology

The data are 39 rank orders. It seems
rather procrustean to approximate  discrete rank numbers by
continuous distances. And moreover the data are conditional, which means that
a rank equal to five for staff member A may not have the same "meaning" as a five for staff member B. Nevertheless we use the data for our example, since it is after all an off-diagonal rectangular matrix with positive numbers.

All rows of $D$ add up to 45, so if we were to choose row-means for $f$ they would
be exactly zero. If we use all of the 9 rows of $F$ in turn, then for
six of them we find that $M$ is indefinite. The configurations for rows 1, 2, and 7 are plotted.

```{r rosdata, echo = FALSE}
roskam <-
structure(list(SOC = c("1", "1", "1", "1", "7", "6", "2", "4", 
"4", "3", "4", "3", "2", "2", "7", "5", "5", "9", "9", "8", "7", 
"8", "8", "8", "7", "4", "5", "1", "2", "2", "5", "4", "5", "6", 
"8", "2", "5", "8", "5"), EDU = c("5", "3", "6", "5", "1", "1", 
"1", "1", "1", "1", "1", "2", "9", "7", "2", "7", "9", "6", "6", 
"3", "2", "7", "6", "7", "3", "7", "6", "8", "5", "5", "3", "5", 
"7", "3", "5", "6", "8", "7", "6"), CLI = c("7", "2", "5", "4", 
"4", "2", "4", "2", "3", "2", "8", "1", "1", "1", "1", "8", "8", 
"5", "7", "7", "8", "6", "5", "5", "6", "9", "8", "9", "6", "4", 
"2", "6", "9", "7", "7", "5", "9", "3", "7"), MAT = c("3", "7", 
"3", "7", "3", "5", "5", "8", "5", "4", "3", "5", "6", "4", "3", 
"1", "1", "1", "2", "2", "5", "3", "2", "2", "2", "5", "2", "2", 
"4", "3", "9", "2", "3", "2", "4", "4", "2", "4", "2"), EXP = c("2", 
"6", "8", "6", "6", "3", "3", "3", "7", "6", "7", "6", "8", "3", 
"5", "3", "2", "3", "1", "1", "1", "1", "1", "1", "1", "1", "1", 
"3", "8", "6", "4", "8", "2", "8", "2", "3", "3", "2", "4"), 
    CUL = c("4", "4", "2", "2", "8", "7", "8", "5", "6", "8", 
    "6", "8", "3", "9", "8", "9", "7", "7", "8", "9", "9", "9", 
    "9", "9", "9", "8", "9", "7", "1", "1", "1", "7", "8", "5", 
    "9", "7", "7", "9", "9"), IND = c("6", "5", "4", "3", "2", 
    "8", "6", "9", "8", "9", "2", "7", "4", "5", "9", "4", "6", 
    "8", "3", "4", "6", "2", "4", "6", "8", "2", "4", "6", "7", 
    "8", "6", "1", "1", "1", "1", "1", "1", "5", "8"), TST = c("9", 
    "8", "7", "8", "9", "4", "7", "6", "2", "7", "5", "4", "5", 
    "6", "4", "2", "3", "2", "4", "5", "4", "5", "7", "4", "4", 
    "3", "7", "4", "3", "7", "7", "3", "4", "4", "3", "8", "4", 
    "6", "3"), PHY = c("8", "9", "9", "9", "5", "9", "9", "7", 
    "9", "5", "9", "9", "7", "8", "6", "6", "4", "4", "5", "6", 
    "3", "4", "3", "3", "5", "6", "3", "5", "9", "9", "8", "9", 
    "6", "9", "6", "9", "6", "1", "1")), row.names = c(NA, 39L
), class = "data.frame")
roskamNumeric <- matrix(as.numeric(as.matrix(roskam)), 39, 9)
roskamColLabels <- colnames(roskam)
roskamObjects <- as.character(apply(roskamNumeric, 1, function(x) which(x == 1)))
h1 <- schoenemann(roskamNumeric, fsel = 1)
h2 <- schoenemann(roskamNumeric, fsel = 2)
h7 <- schoenemann(roskamNumeric, fsel = 7)
```

```{r rosplot1, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Roskam Data, Configuration Plot from Column 1"}
par(pty="s")
plot(rbind(h1$xhat, h1$yhat), type = "n", xlab = "dim 1", ylab = "dim 2")
  text(h1$xhat,
       as.character(1:39),
       col = "RED",
       cex = 1.5)
  text(h1$yhat,
       roskamColLabels,
       col = "BLUE",
       cex = 1.5)
```
```{r rosplot2, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Roskam Data, Configuration Plot from Column 2"}
par(pty="s")
plot(rbind(h2$xhat, h2$yhat), type = "n", xlab = "dim 1", ylab = "dim 2")
  text(h2$xhat,
       as.character(1:39),
       col = "RED",
       cex = 1.5)
  text(h2$yhat,
       roskamColLabels,
       col = "BLUE",
       cex = 1.5)
```
```{r rosplot3, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Roskam Data, Configuration Plot from Column 7"}
par(pty="s")
  plot(rbind(h7$xhat, h7$yhat), type = "n", xlab = "dim 1", ylab = "dim 2")
  text(h7$xhat,
       as.character(1:39),
       col = "RED",
       cex = 1.5)
  text(h7$yhat,
       roskamColLabels,
       col = "BLUE",
       cex = 1.5)
```

```{r rosdd, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Roskam Data, D-DHAT plot, Column 7 Solution"}
plot(h7$d, h7$dhat, xlab = "D", ylab = "DHAT", pch = 19, col = "BLUE", cex = 1.5)
gx <- ifelse(outer(as.vector(h7$d), 1:9, "=="), 1, 0)
mx <- crossprod(gx, as.vector(h7$dhat)) / colSums(gx)
gg <- cbind(1:9, mx)
lines(gg, col = "RED", lwd = 4)
``` 

\sectionbreak

# Software

The function schoenemann() in the file schoenemann.R implements the "row-oriented" solution to perfect or imperfect data. We choose a value
of $p$, with $p=2$ the default. The singular value decomposition is used to approximate $D$
by a matrix of rank $p$. We then compute $\tilde F$. If the parameter
fsel is zero, $f$ are the row-averages of $\tilde F$. If fsel is equal to
$j$ then $f$ is column $j$ of $\tilde F$. The R function lm.fit() is used
to find the least squares solution of the linear system ..., and the
program gives a warning if the estimated rank of $K$ is less than
its number of columns $\frac12 p(p+3)$. From the solution we
reconstruct $M$, and again a warning is issued if $M$ is not 
positive definite. The R function eigen() uses $M=L\Lambda L'$ to test for positive definiteness. If some eigenvalues are negative then
we replace them by their absolute values. There is not real
convincing justification for this, except the lame excuse 
that if $M$, which is of order $p$, only has small negative eigenvalues
then we only make a small unjustified adjustment.  The function 
schoenemann() does return $\tilde F$, $K$, and $M$ so we can 
do some post-hoc analyses to see where things went wrong. In addition
solutions $X$ and $Y$, with corresponding $D(X,Y)$ are returned, as well as the
input matrix $D$.



\sectionbreak

# References